---
title: "Table 7 Lab: Field Dataset Cleaning - Solutions"
format: html
editor: source
author: "Ansita Agrawal, Sage Kruleski, Auster Zhang"
embed-resources: true
code-tools: true
echo: true #always show code
error: true #keep rendering even if errors occur
---

## Preamble

## Load appropriate libraries
```{r}
library(dplyr)
library(tidyverse)
library(knitr)
```

## Loading in files
```{r}
v1 <- read_csv("clean_p_regilla_V1_field_data.csv")
v2 <- read_csv("clean_p_regilla_V2_field_data.csv")
```

## Question 1: Combining and Joining Ecological Data

The files `p_regilla_V1_field_data.csv` and `p_regilla_V2_field_data.csv` were collected using different versions of the same Survey123 form. Before we can analyze the data, we need to clean the column names, make the two datasets use the same variable names, and combine them into one dataset.

***1a. Clean the column names in both datasets***
**Hint:** This will make the column names easier to work with in R.
```{r}
# 1a solution
v1 <- v1 |>
  rename_with(str_to_camel)
v2 <- v2 |>
  rename_with(str_to_camel)
```

***1b. The following code chunk creates the Full dataset. Use one of the dplyr mutating joins to merge the two datasets***

**Note:** The merge will fail due to some columns having incompatible types. Locate which columns causes these issues and typecast them in the **v1** dataset.
```{r}
# 1b fix
v1 <- v1 |>
  mutate(latitudeManualGps = as.double(latitudeManualGps),
         waterTemperatureC = as.double(waterTemperatureC))
```

## Creating full dataset
```{r}
# 1b solution
full <- v2 |>
  full_join(v1, by = join_by(
      globalId, island, site, latitudeManualGps, longitudeManualGps, elevationM,
      sex, totalMassG, bagMassG, waterTemperatureC, x, y,
      collectionDateAndTime == creationDate, 
      sulMm == svlMm, 
      animalMassG == animalWeight, 
      airTemperatureC == airTemperatureC44,
      substrateType == substrate
    )
  )
```

## Question 2: Cleaning Date-Time Variables

The application used to record the data automatically logs the data and time in UTC, however the surveys takes place in Coastal California. Fix the 'collectionDateAndTime' column to match the correct time zone.

```{r}
# Q2 solution
full <- full |>
  mutate(collectionDateAndTime = parse_date_time(collectionDateAndTime,
                                                 orders = c("mdy HMS", "mdy HM")),
  collectionDateAndTime = with_tz(collectionDateAndTime, 
                                   tzone = "America/Los_Angeles"))
```

## Question 5: Temperature conversion (functions)

Come columns measures temperature in Celsius, however for plotting and making the data more understandable, it is common to use Fahrenheit. Create a function that takes in the dataframe and using the columns containing Celsius values as inputs, convert those into Fahrenheit (we are creating a function since there are some columns that will be filled in later, such as dorsal and cloacal, that may want to be converted into the future).
```{r}
# Q5 Solution
celsius_to_fahrenheit <- function(data, cols) {
  data |>
    mutate(across({{cols}}, ~ .x * 9/5 + 32))
}
```

Run the code below to test your function:
```{r}
full <- full |> 
  celsius_to_fahrenheit(c(waterTemperatureF = waterTemperatureC, airTemperatureF = airTemperatureC))
full |>
  select(waterTemperatureC, waterTemperatureF, airTemperatureC, airTemperatureF) |> 
  slice_head(n = 5) |> 
  kable()
```

#Question 8 - make a map
```{r}
#required packages
library(tidyverse)
#install.packages("auk")
library(auk)
#install.packages("rebird")
library(rebird)
#install.packages("sf")
library(sf)
#install.packages("terrainr")
library(terrainr)
```

```{r}

full |> 
  filter(x != 0, y != 0)

frog_geo <- st_as_sf(full,coords=c(x = "x", y = "y"),crs="WGS84") #Converts a regular dataframe → sf (simple features) object

frog_geo_buf <- frog_geo |> st_buffer(dist = 5000)

frog_tiles <- get_tiles(frog_geo_buf, services = "USGSTopo", resolution = 200) #Downloads map imagery (tiles) for your spatial extent

ggplot() +
  geom_spatial_rgb(data = frog_tiles$USGSTopo, mapping = aes(x, y, r = red, g = green, b = blue)) +
  geom_sf(data = frog_geo, alpha = .75) +
  labs(title = "xxx", 
       caption="yyy")


```


#notes for making a table with a summary function for body measurements
```{r}
#step 1 = clean site names
clean_names <- full |>
  mutate(
    site = case_when(
      str_detect(str_to_lower(site), "san antonio") ~ "San Antonio creek
",
      str_detect(str_to_lower(site), "ucsb") ~ "UCSB North Open Space
",
      str_detect(str_to_lower(site), "rattlesnake") ~ "Rattlesnake Canyon Creek
",
str_detect(str_to_lower(site), "rattlesnake") ~ "Rattlesnake Canyon Creek
",
str_detect(str_to_lower(site), "rocky nook") ~ "Rocky Nook Park
",
str_detect(str_to_lower(site), "hazard") ~ "Hazard Creek Beach
",
str_detect(str_to_lower(site), "cascada") ~ "Cascada Canyon
",
str_detect(str_to_lower(site), "prisoners") ~ "Prisoners Harbor
",
str_detect(str_to_lower(site), "valley") ~ "Valley Anchorage Harbor
",
str_detect(str_to_lower(site), "del") ~ "Canyon Del Medio
",
str_detect(str_to_lower(site), "watt") ~ "Watt Road Pond
",
str_detect(str_to_lower(site), "trestle") ~ "Trestle Beach
",
str_detect(str_to_lower(site), "bear") ~ "Bear Creek Pond
",
str_detect(str_to_lower(site), "field") ~ "SCI Field Station
",
str_detect(str_to_lower(site), "1127") ~ "SCI Field Station
",
str_detect(str_to_lower(site), "along creek") ~ "SCI Field Station
",
      TRUE ~ site
    )
  )

```


```{r}

#create mass column, remove sex = UK and unknown so we are only looking at adult frogs
clean_names <- clean_names |> 
  mutate(mass = totalMassG - bagMassG) |> 
 filter(!sex %in% c("UK", "Unknown"))
unique(clean_names$sex)

frog_summary <- function(data, ..., summary_vars) {
  data |>
    group_by(...) |>
    summarise(
      across(
        {{ summary_vars }},
        list(
          mean = ~ mean(.x, na.rm = TRUE),
          sd = ~ sd(.x, na.rm = TRUE),
          min = ~ min(.x, na.rm = TRUE),
          max = ~ max(.x, na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
      ),
      n = n()
    )
}

clean_names |> 
  frog_summary(site, sex, summary_vars = c(sulMm, mass))
```

