---
title: "Table 7 Lab: Field Dataset Cleaning - Solutions"
format: html
editor: source
author: "Ansita Agrawal, Sage Kruleski, Auster Zhang"
embed-resources: true
code-tools: true
echo: true #always show code
error: true #keep rendering even if errors occur
---

## Preamble

## Load appropriate libraries
```{r}
library(dplyr)
library(tidyverse)
library(knitr)
library(maps)
```

## Loading in files
```{r}
v1 <- read_csv("clean_p_regilla_V1_field_data.csv")
v2 <- read_csv("clean_p_regilla_V2_field_data.csv")
```

## Question 1: Combining and Joining Ecological Data

The files `p_regilla_V1_field_data.csv` and `p_regilla_V2_field_data.csv` were collected using different versions of the same Survey123 form. Before we can analyze the data, we need to clean the column names, make the two datasets use the same variable names, and combine them into one dataset.

***1a. Clean the column names in both datasets***
**Hint:** This will make the column names easier to work with in R.
```{r}
# 1a solution
v1 <- v1 |>
  rename_with(str_to_camel)
v2 <- v2 |>
  rename_with(str_to_camel)
```

***1b. The following code chunk creates the Full dataset. Use one of the dplyr mutating joins to merge the two datasets***

**Note:** The merge will fail due to some columns having incompatible types. Locate which columns causes these issues and typecast them in the **v1** dataset.
```{r}
# 1b fix
v1 <- v1 |>
  mutate(latitudeManualGps = as.double(latitudeManualGps),
         waterTemperatureC = as.double(waterTemperatureC))
```

## Creating full dataset
```{r}
# 1b solution
full <- v2 |>
  full_join(v1, by = join_by(
      globalId, island, site, latitudeManualGps, longitudeManualGps, elevationM,
      sex, totalMassG, bagMassG, waterTemperatureC, x, y,
      collectionDateAndTime == creationDate, 
      sulMm == svlMm, 
      animalMassG == animalWeight, 
      airTemperatureC == airTemperatureC44,
      substrateType == substrate
    )
  )
```

## Question 2: Cleaning Date-Time Variables

The application used to record the data automatically logs the data and time in UTC, however the surveys takes place in Coastal California. Fix the 'collectionDateAndTime' column to match the correct time zone.

```{r}
# Q2 solution
full <- full |>
  mutate(collectionDateAndTime = parse_date_time(collectionDateAndTime,
                                                 orders = c("mdy HMS", "mdy HM")),
  collectionDateAndTime = with_tz(collectionDateAndTime, 
                                   tzone = "America/Los_Angeles"))
```

## Question 3: Prepare data frame to make report - ready table

#Question 3a 
clean site names. The names have been provided to you, use stringr syntax to clean row names

```{r}
#step 1 = clean site names
clean_names <- full |>
  mutate(
    site = case_when(
      str_detect(str_to_lower(site), "san antonio") ~ "San Antonio creek",
      str_detect(str_to_lower(site), "ucsb") ~ "UCSB North Open Space",
      str_detect(str_to_lower(site), "rattlesnake") ~ "Rattlesnake Canyon Creek",
str_detect(str_to_lower(site), "rattlesnake") ~ "Rattlesnake Canyon Creek",
str_detect(str_to_lower(site), "rocky nook") ~ "Rocky Nook Park",
str_detect(str_to_lower(site), "hazard") ~ "Hazard Creek Beach",
str_detect(str_to_lower(site), "cascada") ~ "Cascada Canyon",
str_detect(str_to_lower(site), "prisoners") ~ "Prisoners Harbor",
str_detect(str_to_lower(site), "valley") ~ "Valley Anchorage Harbor",
str_detect(str_to_lower(site), "del") ~ "Canyon Del Medio",
str_detect(str_to_lower(site), "watt") ~ "Watt Road Pond",
str_detect(str_to_lower(site), "trestle") ~ "Trestle Beach",
str_detect(str_to_lower(site), "bear") ~ "Bear Creek Pond",
str_detect(str_to_lower(site), "field") ~ "SCI Field Station",
str_detect(str_to_lower(site), "1127") ~ "SCI Field Station",
str_detect(str_to_lower(site), "along creek") ~ "SCI Field Station",
      is.na(site) ~ "SCI Field Station",
      TRUE ~ "SCI Field Station"
    )
  )

```

#Question 3b
calculate animal mass for column 'animalMassG', remove NA and unknown values for 'sex' so we are only looking at adult frogs
```{r}
library(kableExtra)

#calculate animal mass, remove NA and unknown values for 'sex' so we are only looking at adult frogs
clean_names <- clean_names |> 
  mutate(animalMassG = totalMassG - bagMassG) |> 
 filter(!sex %in% c("UK", "Unknown")) |> 
  drop_na(sex)
unique(clean_names$sex)
```


#Question 4: Create function for calculating body measurements by user input columns

#Question 4: 
make a function named frog_summary() that takes a dataframe and calculates measurement values (mean, sd, min, and max), counts the sample size (n()), and renames columns from user-input columns
```{r}

frog_summary <- function(data, ..., summary_vars) {
  data |>
    group_by(...) |>
    summarise(
      across(
        {{ summary_vars }},
        list(
          mean = ~ mean(.x, na.rm = TRUE),
          sd = ~ sd(.x, na.rm = TRUE),
          min = ~ min(.x, na.rm = TRUE),
          max = ~ max(.x, na.rm = TRUE)
        ),
        .names = "{.col}_{.fn}"
      ),
      n = n()
    )
}
```

## Question 5: Body mass measurements/averages table

#Question 5a 
make a prep table by applying function, then ungroup so that the formatted table can include island and mainland as group headers.
```{r}

#prep table by applying function, then ungrouping so table can include island and mainland as group headers
table_data <- clean_names |> 
  frog_summary(island, site, sex, summary_vars = c(sulMm, animalMassG)) |> 
  ungroup() |> 
  arrange(island, site)
```

y#Question 5b
create a table that shows mass and sul as table headers for columns and island and mainland for grouped headers for rows
(show example table, something similar to lab 7)
```{r}
  
#create a table that shows mass and sul as table headers for columns and island and mainland for grouped headers for rows
table <- table_data |> 
  select(-island) |> 
   select(site, sex, n, everything()) |> 
  kable(digits = c(0, 1, 1, 0),
        col.names = c("Site", "Sex", "N", "Mean", "SD", "Min", "Max","Mean", "SD", "Min", "Max" ),
        caption = "Pacific chorus frog measurements. Frogs were captured on Santa Cruz Island and along adjacent mainland California sites.") |>
  kable_classic(full_width = F,
                bootstrap_options = "striped") |> 
  add_header_above(c(" " = 3, "SUL (mm)" = 4, "Mass (g)" = 4),
                   bold = TRUE) |> 
  pack_rows("Mainland",
            1,
            sum(table_data$island == "Mainland")) |> 
  pack_rows("Island",
            sum(table_data$island == "Mainland") + 1,
            nrow(table_data)) |> 
  row_spec(row = 0, bold = T, align = "c") |> 
  footnote(general = "SUL is snout urostyle length")

table
```

## Question 6: Temperature conversion (functions)

Come columns measures temperature in Celsius, however for plotting and making the data more understandable, it is common to use Fahrenheit. Create a function that takes in the dataframe and using the columns containing Celsius values as inputs, convert those into Fahrenheit (we are creating a function since there are some columns that will be filled in later, such as dorsal and cloacal, that may want to be converted into the future).
```{r}
# Q6 Solution
celsius_to_fahrenheit <- function(data, cols) {
  data |>
    mutate(across({{cols}}, ~ .x * 9/5 + 32))
}
```

Run the code below to test your function:
```{r}
full <- full |> 
  celsius_to_fahrenheit(c(waterTemperatureF = waterTemperatureC, airTemperatureF = airTemperatureC))
full |>
  select(waterTemperatureC, waterTemperatureF, airTemperatureC, airTemperatureF) |> 
  slice_head(n = 5) |> 
  kable()
```

## Question 7: Visualizing Frog Body Measurements
Use the cleaned dataset to explore differences in frog body measurements across regions and sexes.

Create a figure that compares body measurements across sexes and regions.

Your figure must:
- use `ggplot2`
- include labels and a title
- use faceting by either `site` or `sex`

```{r}

figure_data <- clean_names |>
  filter(!is.na(sex),
         !is.na(sulMm),
         !is.na(animalMassG),
         !is.na(island)) |>
  group_by(island, site, sex) |>
  summarise(
    avg_sul = mean(sulMm, na.rm = TRUE),
    avg_mass = mean(animalMassG, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

ggplot(figure_data, aes(x = avg_sul, y = avg_mass, color = sex)) +
  geom_point(aes(size = n), alpha = 0.8) +
  facet_wrap(~ island) +
  labs(
    title = "Average Frog Body Size and Mass by Sex and Region",
    subtitle = "Each point represents one site and sex group",
    x = "Average SUL (mm)",
    y = "Average Animal Mass (g)",
    color = "Sex",
    size = "Sample Size"
  ) +
  theme_minimal()
```


## Question 8: Create a map of sampling locations

Using the latitude and longitude GPS data, map the observation on the California map that also shows the county lines. Use the columns 'x' and 'y', not latitude/longitudeManualGPS.

```{r}
# Q8 Solution
# when plotting the GPS points, one observation recorded (0,0), which is not a valid point. Users must get rid of this datapoint somehow. How the map is created is free to the user. An alpha is incorporated so that overlaps of points are explicit
county <- map_data("county")
ca <- filter(county, region == "california")

ggplot(data = ca) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "white", color = "grey25", linewidth = 0.1) +
  coord_map() +
  geom_point(data = filter(full, y > 0), aes(x = x, y = y), color = "steelblue", alpha = 0.1) +
  xlab("") + ylab("") +
  theme(plot.title = element_text(hjust = 0.5),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "none",
        plot.margin = margin(0, 0, 0, 0, "cm"))
```



## References
- [Creating Date/Time objects with inconsistent formats](https://www.r-bloggers.com/2024/09/mastering-date-and-time-data-in-r-with-lubridate/#:~:text=Solving%20Real%2DWorld%20Date%2DTime%20Issues)
- [Renaming Columns Headers](https://www.statology.org/r-rename_with/)
- [Mapping California](https://www.linkedin.com/learning/data-visualization-in-r-with-ggplot2-24583012/zooming-in-on-california)
